cardinalR: Generating interesting high-dimensional data structures

### 2025-09-19 paper-cardinalR.Rmd
Updated trigometric section

Trigonometric-based structures provide flexible ways to simulate complex curved patterns and spirals that often arise in real-world high-dimensional data, such as in biological trajectories, or physical systems (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:triginometric)"} else { "Figure \\@ref(fig:triginometric-proj)"})`). The package includes functions to generate a wide range of these forms from crescents and curvy cylinders to spherical, helical, and conic spirals, as well as nonlinear hyperbolas. These structures are particularly valuable for testing how well dimension reduction and clustering algorithms preserve intricate geometric and topological features. Table `r knitr::asis_output(ifelse(knitr::is_html_output(), '\\@ref(tab:trigonometric-tb-html)', '\\@ref(tab:trigonometric-tb-pdf)'))` summarizes these functions.

@@

#### `gen_crescent()`

The `gen_crescent(n, p)` function generates a $p$-dimensional dataset of $n$ observations based on a $2\text{-}D$ crescent-shaped manifold with optional structured high-dimensional noise (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:triginometric) a"} else { "Figure \\@ref(fig:triginometric-proj) a"})`).

The first two dimensions define a crescent-shaped curve using a semi-circular arc. Let $\theta \in [\pi/6, 2\pi]$ be a sequence of $n$ evenly spaced angles. The corresponding $2\text{-}D$ coordinates are defined by: $X_1 = \cos(\theta)$, $X_2 = \sin(\theta)$.

@@

#### `gen_curvycylinder()`

The `gen_curvycylinder(n, p, h)` function generates a $p$-dimensional dataset of $n$ observations structured as a $3\text{-}D$ cylindrical manifold with an added nonlinear curvy dimension, and optional noise dimensions when $p > 4$ (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:triginometric) b"} else { "Figure \\@ref(fig:triginometric-proj) b"})`).

The core structure consists of a circular base and height values, extended by a nonlinear fourth dimension. Let $\theta \sim U(0, 3\pi)$ represent a random angle on a circular base and $z \sim U(0, h)$ represent the height along the cylinder. The coordinates are defined as: $X_1 = \cos(\theta)$ (Circular base, x-axis), $X_2 = \sin(\theta)$ (Circular base, y-axis), $X_3 = z$ (Linear height), and $X_4 = \sin(z)$ (Nonlinear curvy variation along height). This forms a curvy cylindrical surface in $4\text{-}D$, where the fourth dimension bends periodically along the height axis, resembling a helicoid or twisting wave along the cylinder.

@@

#### `gen_sphericalspiral()`

The `gen_sphericalspiral(n, p, spins)` function simulates a dataset of $n$ observations that form a spiraling path on a spherical surface in the first four dimensions. When extended beyond $4\text{-}D$, structured nonlinear noise dimensions are added to simulate more realistic high-dimensional manifolds (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:triginometric) c"} else { "Figure \\@ref(fig:triginometric-proj) c"})`).

The first three dimensions represent points on a unit sphere. Let $\theta \in [0, 2\pi \times \text{spins}]$ be the azimuthal angle (longitude), controls the number of spiral turns and the $\phi \in [0, \pi]$be the polar angle (latitude), controls the vertical sweep from the north to the south pole.

Cartesian coordinates from spherical conversion: $X_1 = \sin(\phi)\cos(\theta)$, $X_2 = \sin(\phi)\sin(\theta)$, $X_3 = \cos(\phi) + \varepsilon$, where $\varepsilon \sim U(-0.5, 0.5)$ introduces vertical jitter, and $X_4 = \theta / \max(\theta)$: a normalized progression along the spiral path. This generates a spherical spiral curve embedded in $4\text{-}D$ space, combining both circular and vertical movement, with gentle curvature and non-linear progression.

@@

#### `gen_helicalspiral()`

The `gen_helicalspiral(n, p)` function generates a dataset of $n$ observations forming a helical spiral, embedded in the first four dimensions of $p\text{-}D$. Additional dimensions (if $p > 4$) are filled with structured noise to mimic high-dimensional complexity (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:triginometric) d"} else { "Figure \\@ref(fig:triginometric-proj) d"})`).

The first four coordinates follow a $3\text{-}D$ helix with an additional oscillating component. Let $\theta \in [0, 5\pi/4]$ be a sequence of angles controlling rotation around a circle. Cartesian coordinates; $X_1 = \cos(\theta)$: circular trajectory along the x-axis, $X_2 = \sin(\theta)$: circular trajectory along the y-axis, $X_3 = 0.05\theta + \varepsilon_3$, with $\varepsilon_3 \sim U(-0.5, 0.5)$: linear progression (height) with vertical jitter, simulating a helix, and $X_4 = 0.1\sin(\theta)$: oscillates with $\theta$, representing a periodic "wobble" along the fourth dimension. This results in a helical spiral winding upward along the third axis, with gentle sinusoidal fluctuation in the fourth dimension.

@@

#### `gen_conicspiral()`

The `gen_conicspiral(n, p, spins)` function generates a dataset of $n$ points forming a conical spiral in the first four dimensions of $p\text{-}D$ (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:triginometric) e"} else { "Figure \\@ref(fig:triginometric-proj) e"})`). The geometry combines radial expansion, vertical elevation, and spiral deformation, simulating a structure that fans out like a $3\text{-}D$ conic helix.

The shape is defined by parameter $\theta \in [0, 2\pi\text{spins}]$, controlling the angular progression of the spiral. The Archimedean spiral in the horizontal plane is represented by; $X_1 = \theta\cos(\theta)$ for radial expansion in x, and $X_2 = \theta\sin(\theta)$ for radial expansion in y. The growth pattern resembles a cone, with the height increasing according to $X_3 = 2\theta / \max(\theta) + \varepsilon_3$, with $\varepsilon_3 \sim U(-0.1, 0.6).$ Spiral modulation in the fourth dimension is represented by $X_4 = \theta\sin(2\theta) + \varepsilon_4$, with $\varepsilon_4 \sim U(-0.1, 0.6)$ which simulates a twisting helical component in a non-radial dimension. This results in a $3\text{-}D$ spiral surface expanding upward and outward like a cone, with an oscillatory fourth dimension capturing spiral irregularities.

@@

#### `gen_nonlinear()`

The `gen_nonlinear(n, p, hc, non_fac)` function simulates a non-linear $2\text{-}D$ surface embedded in higher dimensions, constructed using inverse and trigonometric transformations applied to independent variables (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:triginometric) f"} else { "Figure \\@ref(fig:triginometric-proj) f"})`).

The $X_{1} \sim U(0.1, 2)$: base variable (avoids zero to prevent division errors), $X_{3} \sim U(0.1, 0.8)$: independent auxiliary variable, $X_{2} = hc/X_{1} + \text{nonfac}\sin(X_{1})$: non-linear combination of hyperbolic and sinusoidal transformations, creating sharp curvature and oscillation, and $X_{4} = \cos(\pi X_{1}) + \varepsilon$, with $\varepsilon \sim U(-0.1, 0.1)$: additional nonlinear variation based on cosine, simulating more subtle periodic structure. These transformations together result in a non-linear surface warped in multiple ways: sharp vertical shifts due to inverse terms, smooth waves from sine and cosine, and additional jitter.

@@

For trigonometric shapes, additional noise dimensions are appended to embed the structure into higher-dimensional space: for the crescent-shaped curve this applies when $p > 2$ (adding $X_3$ to $X_p$), while for all other trigonometric shapes it applies when $p > 4$ (adding $X_5$ to $X_p$).

>>>>

Trigonometric-based structures provide flexible ways to simulate complex curved patterns and spirals that often arise in real-world high-dimensional data, such as in biological trajectories, or physical systems (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:triginometric)"} else { "Figure \\@ref(fig:triginometric-proj)"})`). The first few coordinates define the main geometric structure (e.g., crescent, cylinder, spiral, or helix), while for higher-dimensional embeddings, additional noise dimensions are appended: for the crescent shape this occurs when $p > 2$ (adding $X_3$ through $X_p$), and for all other trigonometric shapes when $p > 4$ (adding $X_5$ through $X_p$). These structures are particularly valuable for testing how well dimension reduction and clustering algorithms preserve intricate geometric and topological features. Table `r knitr::asis_output(ifelse(knitr::is_html_output(), '\\@ref(tab:trigonometric-tb-html)', '\\@ref(tab:trigonometric-tb-pdf)'))` summarizes these functions.

@@

First, the `gen_crescent(n, p)` function generates a $p$-dimensional dataset of $n$ observations based on a $2\text{-}D$ crescent-shaped manifold with optional structured high-dimensional noise (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:triginometric) a"} else { "Figure \\@ref(fig:triginometric-proj) a"})`). Let $\theta \in [\pi/6, 2\pi]$ be a sequence of $n$ evenly spaced angles. The corresponding $2\text{-}D$ coordinates are defined by: $X_1 = \cos(\theta)$, $X_2 = \sin(\theta)$.

@@

Second, the `gen_curvycylinder(n, p, h)` function generates a $p$-dimensional dataset of $n$ observations structured as a $3\text{-}D$ cylindrical manifold with an added nonlinear curvy dimension, and optional noise dimensions when $p > 4$ (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:triginometric) b"} else { "Figure \\@ref(fig:triginometric-proj) b"})`). The core structure consists of a circular base and height values, extended by a nonlinear fourth dimension. Let $\theta \sim U(0, 3\pi)$ represent a random angle on a circular base and $z \sim U(0, h)$ represent the height along the cylinder. The coordinates are defined as: $X_1 = \cos(\theta)$ (Circular base, x-axis), $X_2 = \sin(\theta)$ (Circular base, y-axis), $X_3 = z$ (Linear height), and $X_4 = \sin(z)$ (Nonlinear curvy variation along height).

@@

For a spiraling path on a spherical surface in the first four dimensions, `gen_sphericalspiral(n, p, spins)` (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:triginometric) c"} else { "Figure \\@ref(fig:triginometric-proj) c"})`), let $\theta \in [0, 2\pi \times \text{spins}]$ be the azimuthal angle (longitude), controls the number of spiral turns and the $\phi \in [0, \pi]$be the polar angle (latitude), controls the vertical sweep from the north to the south pole. Cartesian coordinates from spherical conversion: $X_1 = \sin(\phi)\cos(\theta)$, $X_2 = \sin(\phi)\sin(\theta)$, $X_3 = \cos(\phi) + \varepsilon$, where $\varepsilon \sim U(-0.5, 0.5)$ introduces vertical jitter, and $X_4 = \theta / \max(\theta)$: a normalized progression along the spiral path. This generates a spherical spiral curve embedded in $4\text{-}D$ space, combining both circular and vertical movement, with gentle curvature and non-linear progression.

@@

For a helical spiral in four dimensions, `gen_helicalspiral(n, p)` (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:triginometric) d"} else { "Figure \\@ref(fig:triginometric-proj) d"})`), let $\theta \in [0, 5\pi/4]$ be a sequence of angles controlling rotation around a circle. Cartesian coordinates; $X_1 = \cos(\theta)$: circular trajectory along the x-axis, $X_2 = \sin(\theta)$: circular trajectory along the y-axis, $X_3 = 0.05\theta + \varepsilon_3$, with $\varepsilon_3 \sim U(-0.5, 0.5)$: linear progression (height) with vertical jitter, simulating a helix, and $X_4 = 0.1\sin(\theta)$: oscillates with $\theta$, representing a periodic "wobble" along the fourth dimension.

@@

Similarly, the `gen_conicspiral(n, p, spins)` function generates a dataset of $n$ points forming a conical spiral in the first four dimensions of $p\text{-}D$ (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:triginometric) e"} else { "Figure \\@ref(fig:triginometric-proj) e"})`). The geometry combines radial expansion, vertical elevation, and spiral deformation, simulating a structure that fans out like a $3\text{-}D$ conic helix. The shape is defined by parameter $\theta \in [0, 2\pi\text{spins}]$, controlling the angular progression of the spiral. The Archimedean spiral in the horizontal plane is represented by; $X_1 = \theta\cos(\theta)$ for radial expansion in x, and $X_2 = \theta\sin(\theta)$ for radial expansion in y. The growth pattern resembles a cone, with the height increasing according to $X_3 = 2\theta / \max(\theta) + \varepsilon_3$, with $\varepsilon_3 \sim U(-0.1, 0.6).$ Spiral modulation in the fourth dimension is represented by $X_4 = \theta\sin(2\theta) + \varepsilon_4$, with $\varepsilon_4 \sim U(-0.1, 0.6)$ which simulates a twisting helical component in a non-radial dimension.

@@

Finally, the `gen_nonlinear(n, p, hc, non_fac)` function simulates a non-linear $2\text{-}D$ surface embedded in higher dimensions, constructed using inverse and trigonometric transformations applied to independent variables (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:triginometric) f"} else { "Figure \\@ref(fig:triginometric-proj) f"})`). The $X_{1} \sim U(0.1, 2)$: base variable (avoids zero to prevent division errors), $X_{3} \sim U(0.1, 0.8)$: independent auxiliary variable, $X_{2} = hc/X_{1} + \text{nonfac}\sin(X_{1})$: non-linear combination of hyperbolic and sinusoidal transformations, creating sharp curvature and oscillation, and $X_{4} = \cos(\pi X_{1}) + \varepsilon$, with $\varepsilon \sim U(-0.1, 0.1)$: additional nonlinear variation based on cosine, simulating more subtle periodic structure. These transformations together result in a non-linear surface warped in multiple ways: sharp vertical shifts due to inverse terms, smooth waves from sine and cosine, and additional jitter.

@@

### 2025-09-19 paper-cardinalR.tex
Updated trigometric section

Trigonometric-based structures provide flexible ways to simulate complex curved patterns and spirals that often arise in real-world high-dimensional data, such as in biological trajectories, or physical systems (Figure \ref{fig:triginometric-proj}). The package includes functions to generate a wide range of these forms from crescents and curvy cylinders to spherical, helical, and conic spirals, as well as nonlinear hyperbolas. These structures are particularly valuable for testing how well dimension reduction and clustering algorithms preserve intricate geometric and topological features. Table \ref{tab:trigonometric-tb-pdf} summarizes these functions.

@@

\paragraph{\texorpdfstring{\texttt{gen\_crescent()}}{gen\_crescent()}}\label{gen_crescent}

The \texttt{gen\_crescent(n,\ p)} function generates a \(p\)-dimensional dataset of \(n\) observations based on a \(2\text{-}D\) crescent-shaped manifold with optional structured high-dimensional noise (Figure \ref{fig:triginometric-proj} a).

The first two dimensions define a crescent-shaped curve using a semi-circular arc. Let \(\theta \in [\pi/6, 2\pi]\) be a sequence of \(n\) evenly spaced angles. The corresponding \(2\text{-}D\) coordinates are defined by: \(X_1 = \cos(\theta)\), \(X_2 = \sin(\theta)\).

@@

\paragraph{\texorpdfstring{\texttt{gen\_curvycylinder()}}{gen\_curvycylinder()}}\label{gen_curvycylinder}

The \texttt{gen\_curvycylinder(n,\ p,\ h)} function generates a \(p\)-dimensional dataset of \(n\) observations structured as a \(3\text{-}D\) cylindrical manifold with an added nonlinear curvy dimension, and optional noise dimensions when \(p > 4\) (Figure \ref{fig:triginometric-proj} b).

The core structure consists of a circular base and height values, extended by a nonlinear fourth dimension. Let \(\theta \sim U(0, 3\pi)\) represent a random angle on a circular base and \(z \sim U(0, h)\) represent the height along the cylinder. The coordinates are defined as: \(X_1 = \cos(\theta)\) (Circular base, x-axis), \(X_2 = \sin(\theta)\) (Circular base, y-axis), \(X_3 = z\) (Linear height), and \(X_4 = \sin(z)\) (Nonlinear curvy variation along height). This forms a curvy cylindrical surface in \(4\text{-}D\), where the fourth dimension bends periodically along the height axis, resembling a helicoid or twisting wave along the cylinder.

@@

\paragraph{\texorpdfstring{\texttt{gen\_sphericalspiral()}}{gen\_sphericalspiral()}}\label{gen_sphericalspiral}

The \texttt{gen\_sphericalspiral(n,\ p,\ spins)} function simulates a dataset of \(n\) observations that form a spiraling path on a spherical surface in the first four dimensions. When extended beyond \(4\text{-}D\), structured nonlinear noise dimensions are added to simulate more realistic high-dimensional manifolds (Figure \ref{fig:triginometric-proj} c).

The first three dimensions represent points on a unit sphere. Let \(\theta \in [0, 2\pi \times \text{spins}]\) be the azimuthal angle (longitude), controls the number of spiral turns and the \(\phi \in [0, \pi]\)be the polar angle (latitude), controls the vertical sweep from the north to the south pole.

Cartesian coordinates from spherical conversion: \(X_1 = \sin(\phi)\cos(\theta)\), \(X_2 = \sin(\phi)\sin(\theta)\), \(X_3 = \cos(\phi) + \varepsilon\), where \(\varepsilon \sim U(-0.5, 0.5)\) introduces vertical jitter, and \(X_4 = \theta / \max(\theta)\): a normalized progression along the spiral path. This generates a spherical spiral curve embedded in \(4\text{-}D\) space, combining both circular and vertical movement, with gentle curvature and non-linear progression.

@@

\paragraph{\texorpdfstring{\texttt{gen\_helicalspiral()}}{gen\_helicalspiral()}}\label{gen_helicalspiral}

The \texttt{gen\_helicalspiral(n,\ p)} function generates a dataset of \(n\) observations forming a helical spiral, embedded in the first four dimensions of \(p\text{-}D\). Additional dimensions (if \(p > 4\)) are filled with structured noise to mimic high-dimensional complexity (Figure \ref{fig:triginometric-proj} d).

The first four coordinates follow a \(3\text{-}D\) helix with an additional oscillating component. Let \(\theta \in [0, 5\pi/4]\) be a sequence of angles controlling rotation around a circle. Cartesian coordinates; \(X_1 = \cos(\theta)\): circular trajectory along the x-axis, \(X_2 = \sin(\theta)\): circular trajectory along the y-axis, \(X_3 = 0.05\theta + \varepsilon_3\), with \(\varepsilon_3 \sim U(-0.5, 0.5)\): linear progression (height) with vertical jitter, simulating a helix, and \(X_4 = 0.1\sin(\theta)\): oscillates with \(\theta\), representing a periodic ``wobble'' along the fourth dimension. This results in a helical spiral winding upward along the third axis, with gentle sinusoidal fluctuation in the fourth dimension.

@@

\paragraph{\texorpdfstring{\texttt{gen\_conicspiral()}}{gen\_conicspiral()}}\label{gen_conicspiral}

The \texttt{gen\_conicspiral(n,\ p,\ spins)} function generates a dataset of \(n\) points forming a conical spiral in the first four dimensions of \(p\text{-}D\) (Figure \ref{fig:triginometric-proj} e). The geometry combines radial expansion, vertical elevation, and spiral deformation, simulating a structure that fans out like a \(3\text{-}D\) conic helix.

The shape is defined by parameter \(\theta \in [0, 2\pi\text{spins}]\), controlling the angular progression of the spiral. The Archimedean spiral in the horizontal plane is represented by; \(X_1 = \theta\cos(\theta)\) for radial expansion in x, and \(X_2 = \theta\sin(\theta)\) for radial expansion in y. The growth pattern resembles a cone, with the height increasing according to \(X_3 = 2\theta / \max(\theta) + \varepsilon_3\), with \(\varepsilon_3 \sim U(-0.1, 0.6).\) Spiral modulation in the fourth dimension is represented by \(X_4 = \theta\sin(2\theta) + \varepsilon_4\), with \(\varepsilon_4 \sim U(-0.1, 0.6)\) which simulates a twisting helical component in a non-radial dimension. This results in a \(3\text{-}D\) spiral surface expanding upward and outward like a cone, with an oscillatory fourth dimension capturing spiral irregularities.

@@

\paragraph{\texorpdfstring{\texttt{gen\_nonlinear()}}{gen\_nonlinear()}}\label{gen_nonlinear}

The \texttt{gen\_nonlinear(n,\ p,\ hc,\ non\_fac)} function simulates a non-linear \(2\text{-}D\) surface embedded in higher dimensions, constructed using inverse and trigonometric transformations applied to independent variables (Figure \ref{fig:triginometric-proj} f).

The \(X_{1} \sim U(0.1, 2)\): base variable (avoids zero to prevent division errors), \(X_{3} \sim U(0.1, 0.8)\): independent auxiliary variable, \(X_{2} = hc/X_{1} + \text{nonfac}\sin(X_{1})\): non-linear combination of hyperbolic and sinusoidal transformations, creating sharp curvature and oscillation, and \(X_{4} = \cos(\pi X_{1}) + \varepsilon\), with \(\varepsilon \sim U(-0.1, 0.1)\): additional nonlinear variation based on cosine, simulating more subtle periodic structure. These transformations together result in a non-linear surface warped in multiple ways: sharp vertical shifts due to inverse terms, smooth waves from sine and cosine, and additional jitter.

@@

For trigonometric shapes, additional noise dimensions are appended to embed the structure into higher-dimensional space: for the crescent-shaped curve this applies when \(p > 2\) (adding \(X_3\) to \(X_p\)), while for all other trigonometric shapes it applies when \(p > 4\) (adding \(X_5\) to \(X_p\)).

>>>>

Trigonometric-based structures provide flexible ways to simulate complex curved patterns and spirals that often arise in real-world high-dimensional data, such as in biological trajectories, or physical systems (Figure \ref{fig:triginometric-proj}). The first few coordinates define the main geometric structure (e.g., crescent, cylinder, spiral, or helix), while for higher-dimensional embeddings, additional noise dimensions are appended: for the crescent shape this occurs when \(p > 2\) (adding \(X_3\) through \(X_p\)), and for all other trigonometric shapes when \(p > 4\) (adding \(X_5\) through \(X_p\)). These structures are particularly valuable for testing how well dimension reduction and clustering algorithms preserve intricate geometric and topological features. Table \ref{tab:trigonometric-tb-pdf} summarizes these functions.

@@

First, the \texttt{gen\_crescent(n,\ p)} function generates a \(p\)-dimensional dataset of \(n\) observations based on a \(2\text{-}D\) crescent-shaped manifold with optional structured high-dimensional noise (Figure \ref{fig:triginometric-proj} a). Let \(\theta \in [\pi/6, 2\pi]\) be a sequence of \(n\) evenly spaced angles. The corresponding \(2\text{-}D\) coordinates are defined by: \(X_1 = \cos(\theta)\), \(X_2 = \sin(\theta)\).

@@

Second, the \texttt{gen\_curvycylinder(n,\ p,\ h)} function generates a \(p\)-dimensional dataset of \(n\) observations structured as a \(3\text{-}D\) cylindrical manifold with an added nonlinear curvy dimension, and optional noise dimensions when \(p > 4\) (Figure \ref{fig:triginometric-proj} b). The core structure consists of a circular base and height values, extended by a nonlinear fourth dimension. Let \(\theta \sim U(0, 3\pi)\) represent a random angle on a circular base and \(z \sim U(0, h)\) represent the height along the cylinder. The coordinates are defined as: \(X_1 = \cos(\theta)\) (Circular base, x-axis), \(X_2 = \sin(\theta)\) (Circular base, y-axis), \(X_3 = z\) (Linear height), and \(X_4 = \sin(z)\) (Nonlinear curvy variation along height).

@@

For a spiraling path on a spherical surface in the first four dimensions, \texttt{gen\_sphericalspiral(n,\ p,\ spins)} (Figure \ref{fig:triginometric-proj} c), let \(\theta \in [0, 2\pi \times \text{spins}]\) be the azimuthal angle (longitude), controls the number of spiral turns and the \(\phi \in [0, \pi]\)be the polar angle (latitude), controls the vertical sweep from the north to the south pole. Cartesian coordinates from spherical conversion: \(X_1 = \sin(\phi)\cos(\theta)\), \(X_2 = \sin(\phi)\sin(\theta)\), \(X_3 = \cos(\phi) + \varepsilon\), where \(\varepsilon \sim U(-0.5, 0.5)\) introduces vertical jitter, and \(X_4 = \theta / \max(\theta)\): a normalized progression along the spiral path. This generates a spherical spiral curve embedded in \(4\text{-}D\) space, combining both circular and vertical movement, with gentle curvature and non-linear progression.

@@

For a helical spiral in four dimensions, \texttt{gen\_helicalspiral(n,\ p)} (Figure \ref{fig:triginometric-proj} d), let \(\theta \in [0, 5\pi/4]\) be a sequence of angles controlling rotation around a circle. Cartesian coordinates; \(X_1 = \cos(\theta)\): circular trajectory along the x-axis, \(X_2 = \sin(\theta)\): circular trajectory along the y-axis, \(X_3 = 0.05\theta + \varepsilon_3\), with \(\varepsilon_3 \sim U(-0.5, 0.5)\): linear progression (height) with vertical jitter, simulating a helix, and \(X_4 = 0.1\sin(\theta)\): oscillates with \(\theta\), representing a periodic ``wobble'' along the fourth dimension.

@@

Similarly, the \texttt{gen\_conicspiral(n,\ p,\ spins)} function generates a dataset of \(n\) points forming a conical spiral in the first four dimensions of \(p\text{-}D\) (Figure \ref{fig:triginometric-proj} e). The geometry combines radial expansion, vertical elevation, and spiral deformation, simulating a structure that fans out like a \(3\text{-}D\) conic helix. The shape is defined by parameter \(\theta \in [0, 2\pi\text{spins}]\), controlling the angular progression of the spiral. The Archimedean spiral in the horizontal plane is represented by; \(X_1 = \theta\cos(\theta)\) for radial expansion in x, and \(X_2 = \theta\sin(\theta)\) for radial expansion in y. The growth pattern resembles a cone, with the height increasing according to \(X_3 = 2\theta / \max(\theta) + \varepsilon_3\), with \(\varepsilon_3 \sim U(-0.1, 0.6).\) Spiral modulation in the fourth dimension is represented by \(X_4 = \theta\sin(2\theta) + \varepsilon_4\), with \(\varepsilon_4 \sim U(-0.1, 0.6)\) which simulates a twisting helical component in a non-radial dimension.

@@

Finally, the \texttt{gen\_nonlinear(n,\ p,\ hc,\ non\_fac)} function simulates a non-linear \(2\text{-}D\) surface embedded in higher dimensions, constructed using inverse and trigonometric transformations applied to independent variables (Figure \ref{fig:triginometric-proj} f). The \(X_{1} \sim U(0.1, 2)\): base variable (avoids zero to prevent division errors), \(X_{3} \sim U(0.1, 0.8)\): independent auxiliary variable, \(X_{2} = hc/X_{1} + \text{nonfac}\sin(X_{1})\): non-linear combination of hyperbolic and sinusoidal transformations, creating sharp curvature and oscillation, and \(X_{4} = \cos(\pi X_{1}) + \varepsilon\), with \(\varepsilon \sim U(-0.1, 0.1)\): additional nonlinear variation based on cosine, simulating more subtle periodic structure. These transformations together result in a non-linear surface warped in multiple ways: sharp vertical shifts due to inverse terms, smooth waves from sine and cosine, and additional jitter.

@@
